Title

### 2012-01-20 inc/laser/laser.tex
ported oe refs to #21 inclusive

%TODO is this first sentance correct or should I use "I" voice?
In this section, a high-power, femtosecond Yb:KGW oscillator is presented that also accesses the broader emission bandwidth parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis), yet employs a simpler geometry for polarized diode-pumping on the strongest 980nm absorption line parallel to the optical $N_m$-axis (rotated by $\sim\,20^\circ$ from the crystallographic $a$-axis\cite{Biswal_thermo_optical_05,mochalov_laser_1997,pujol_crystalline_1999}).
The laser head design is based on the thermal lens shaping (TLS) technique developed in-house for astigmatism compensation in diode-pumped Nd-doped lasers with Brewster-cut gain media [16].
It employs a simple single layer \ce{SiO2} anti-reflection coating on the gain medium to provide minimal loss for $p$-polarized intracavity 1040nm laser radiation and less than 0.5\% surface reflection for orthogonal $s$-polarized 980nm pump light.
The result is a simple and robust, soliton mode-locked and directly diode-pumped, solid-state laser oscillator delivering over 4W of average output power in a 63MHz train of pulses with a duration less than 300fs.
Frequency doubling of this laser output in a 2mm Brewster-cut Lithium triborate (LBO) crystal provides 1.65W of average power at 520nm.

The Watt-level green output power from the frequency-doubled sub-picosecond Yb:KGW oscillator corresponds to a visible peak laser pulse power in excess of 100kW, which is well-suited for further frequency conversion through harmonic generation or pumping of an ultrashort pulse optical parametric oscillator.
%TODO change voice/check ->
To our knowledge, the generation of subpicosecond green peak pulse powers greater than 100kW by frequency doubling the output of a laser oscillator has only been demonstrated for high-power thin-disk [17,18] and cavity-dumped [19] Yb-doped solid-state lasers, although externally-doubled commercial Ytterbium femtosecond oscillators\cite{website_amplitude,website_solar} [20,21] and oscillators with rod-like Yb-fiber gain media [22] can be expected to yield comparable peak green pulse powers.
%TODO replace to "This output will then be use to ..." and link to something from previous
%In our case, the visible femtosecond pulses will be used as the driving laser radiation source for an ultrafast electron microscope [23] – a technology that seeks to combine the sub-nanometer spatial resolution of electron microscopy with the sub-picosecond temporal resolution afforded by today’s ultrashort pulse laser systems.
%Critical to this technology is likely to be the development of a femtosecond photo-electron gun with a low spatial emittance [25].
%The coherent control of photoemission through visible laser-driven plasmon excitations on large-area (a few mm2), gold or silver, nano-patterned photocathodes may provide a mechanism by which this can be achieved [25].

@@

The design of the laser head follows the example of Holtom\cite{Holtom_mode_locked_2006}, which ensures efficient pumping and access to the broadest emission bandwidth of Yb:KGW, while also employing the thermal lens shaping (TLS) technique to compensate for astigmatism [16].
To ensure efficient operation of a Yb:KGW laser, the dominant absorption feature at 980nm for radiation polarized parallel to the optical $N_m$-axis ($\sim\,20^\circ$ from the crystallographic a-axis) should be pumped using high brightness diode lasers.
On the other hand, as with \ce{Yb$:$KY(WO4)2} \cite{Liu_diode_pumped_2001,Killi_high_peak_2005}, the broadest emission bandwidth for ultrashort pulse generation occurs for radiation polarized parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis\cite{Holtom_mode_locked_2006}) in this biaxial material. 
%page 8632
Instead of using a polarization-coupled pumping geometry\cite{Holtom_mode_locked_2006} to ensure these optimum pump-lasing conditions, we employ the non-Brewster crystal geometry and cut shown in Fig. 1(a), where a single 193nm-thick \ce{SiO2} anti-reflection coating is applied to both polished 3x10mm Yb:KGW crystal faces.
At an angle of incidence of 32.3$^\circ$, this coating theoretically has zero reflection loss for $p$-polarized 1040nm laser radiation parallel to the $b$-axis propagating perpendicular to the $a$-$b$ plane of the crystal (Fig. 1(b)); i.e., the preferred optical $N_p$-axis with refractive index $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}.
The \ce{SiO2} coating also has an appropriate bandwidth for femtosecond laser operation in excess of 100nm (to the
0.1\% reflection points) for $p$-polarized radiation centered at 1040nm.
On the other hand, the $\sim$ 0.4\% reflection loss per surface for $s$-polarized 1040nm radiation is sufficient to suppress $s$-polarized oscillation in a laser resonator.
In addition, the same anti-reflection coating has less than 0.5\% reflection loss for $s$-polarized 980nm pump radiation (polarized parallel to the $a$-axis of the Yb:KGW crystal) at incidence angles between 21 and 39$^\circ$ (Fig. 1(b)).
Thus, in this Yb:KGW laser crystal geometry, the largest absorption cross-section associated with the optical $N_m$-axis is accessed with about 94\% efficiency ($\cos 20^\circ = 0.94$) by the $s$-polarized pump radiation.

@@

To ensure a stigmatic laser resonator, we also employ the thermal lens shaping (TLS) technique [16] to compensate for the astigmatism induced by the non-normal incidence geometry of the laser crystal. 
For the employed non-Brewster crystal cut (Fig. 1(a)), a ray
%page 8633
transfer matrix analysis (with the $x$ direction parallel to the crystallographic $b$-axis and the $y$
direction parallel to the $a$-axis) reveals that the required ellipticity for a stigmatic thermal lens
generated by a parabolic temperature distribution, $T(x, y) = T_0 - \frac{1}{2} ( A_\smallT x^2 + B_\smallT y^2 )$, where $T_0$
is the on-axis temperature, is given by
%TODO check W symbol consistent with the rest of the paper
\begin{equation} \label{eq:laser_ratio_ba}
  \frac{B_\smallT}{A_\smallT} = 
  \frac{ \cos^2 \theta_2 \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) }
       { \cos^2 \theta_1 \left[ \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) \right] }
  \quad\text{.}
\end{equation}
Here, $\frac{dn}{dT}$ and $n$ are the thermo-optic coefficient and refractive index of the medium respectively (in this case for radiation polarized along the optical $N_p$-axis), $\theta_1$ is the 32.3$^\circ$ angle of incidence, $\theta_2$ = 15.7$^\circ$ is the angle of refraction (as $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}), and $\Delta$ is the fraction of the crystal length $l$ contributing to the bowing of the crystal faces under thermal
expansion with coefficient $\alpha T$ [16].
Generally, $\Delta \approx W l$ , where $W$ is the transverse size (radius) of the pumped region [30].
In our case, $l \approx$ 3mm and $W \approx$ 0.2mm, so that $\Delta < 0.1$, implying that the thermal duct due to a non-zero $\frac{dn}{dT}$ primarily determines the required $B_T$/$A_T$ ratio.
In other words, for the potassium gadolinium tungstate (\ce{KGd(WO4)2}) crystal geometry of Fig. 1(a), we expect from \ref{eq:laser_ratio_ba} to require $\dfrac{B_\smallT}{A_\smallT} \approx \dfrac{ \cos^2 \theta_2 }{ \cos^2 \theta_1 } = 1.3$.

The spatial pump distribution needed to generate the required $B_T/A_T$ ratio can be found using a cosine Fourier series solution to the thermal diffusion equation for our two-dimensional problem,
\begin{equation} \label{eq:heat_eqn}
  \kappa \nabla^2 T(x,y) = Q(x,y)
  \quad\text{,}
\end{equation}
where $\kappa$ is the thermal diffusion coefficient.
Thermal diffusion in the third $z$-direction has been neglected due to the relatively uniform longitudinal heat deposition resulting from our counter-propagating pump geometry [16].
The specifications of our dual-axis collimated diodes [26] indicate that heat source can be written as
\begin{equation}
  Q(x,y) = Q_0 \exp \left[ - \frac{x^2}{W_x^2} - \left( \frac{y^2}{W_y^2} \right)^3 \right]
  \quad\text{;}
\end{equation}
that is, the focused pump radiation is described accurately by a Gaussian irradiance distribution in the horizontal ($x$) direction with half-width 1/e maximum (HW1/eM) spot size $W_x$ and a super-Gaussian of order 3 in the vertical ($y$) direction with spot size $W_y$.
\ref{eq:heat_eqn} can then be solved using the appropriate boundary conditions for our 3x10mm crystal cross-section for the employed TLS technique [16]; namely, heat removal only through the top crystal face implies $T(x,y = \pm 1.5\text{mm}) = 0$ (or arbitrary constant) and $\dfrac{\partial T}{\partial x} = 0$ at $ x = \pm 5.0$mm for any $y$.
For a fixed vertical pump spot size of $W_y = 0.2$mm, our analysis indicates that in order to achieve the required ratio $B_T / A_T = 1.3$ in the parabolic approximation for $T(x,y)$ about $(x,y) = (0,0)$ a horizontal pump spot size $W_x = 0.3$mm is needed; in other words, the required $Q(x,y)$ heat source spot size ratio (or ellipticity) $W_x / W_y = 1.5$.
This analytical result for the central region around the laser beam axis is shown in Fig. 2; displayed are the
%page 8634
horizontal ($y = 0$, Fig. 2(a)) and vertical ($x = 0$, Fig. 2(b)) sections, respectively, through the $Q(x,y)$ and $T(x,y)$ distributions.
Also shown are the parabolic approximations to the pump-induced temperature distribution (dashed lines) and representative TEM$_{00}$ Gaussian laser modes (shaded area) in the Yb:KGW crystal with a HW1/eM field mode sizes of $w_y = 160\mu$m and $w_x = 182 \mu$m (14\% larger due to refraction (Fig. 1(a)).
This mode size ensures efficient pump-probe overlap while minimizing the loss due to ground state absorption in the unpumped regions of the Yb-doped gain medium\cite{Brenier_new_criteria}; in particular, in the vertical direction (Fig. 2(b)) where $w_y$ greater than about $0.8 W_y$ leads to significant absorptive loss as the extrema of the laser mode extend beyond the sharply-bounded super-Gaussian pump region.

@@

We note that the above thermal lens analysis neglects the anisotropy of the thermal conductivity $\kappa$ in potassium gadolinium tungstate (\ce{KGd(WO4)2}) \cite{Biswal_thermo_optical_05,mochalov_laser_1997} and stress-induced refractive index changes [31].
For the employed crystal cut in our TLS geometry (Fig. 1(a)), the anisotropy in $\kappa$ \cite{mochalov_laser_1997} [27] is not expected to have a significant effect; the laser head geometry dictates that the dominant pump-induced heat conduction will be in the saggital ($y$) direction along the crystal $a$-axis where $\kappa \approx 2.6$W/(mK) (heat extraction only through the top and bottom 3x10mm crystal faces), so that the 46\% larger value of $\kappa$ along the $b$-axis ($x$) direction at 3.8W/(mK) should not have a large influence on the heat conduction.
The refractive index
%page 8635
changes due to stress in the pumped gain medium are difficult to quantify since the substantial variations in the reported values of the thermo-optic coefficients in KGW make it impossible to extract either the sign or magnitude of the stress dependence of the refractive index [21].
Moreover, nearly all analyses of thermal lensing effects Nd:KGW and Yb:KGW lasers have been performed for laser radiation polarized along the optical $N_m$-axis, which accesses the largest emission cross-section for both Nd- and Yb-doping [31,34].
Consequently, for our laser polarized along the optical $N_p$-axis\cite{Holtom_mode_locked_2006}, there is insufficient information to include accurately the effects of stress-induced refractive index changes.

>>>>

%TODO is this first sentance correct or should I use "I" voice?
In this section, a high-power, femtosecond Yb:KGW oscillator is presented that also accesses the broader emission bandwidth parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis), yet employs a simpler geometry for polarized diode-pumping on the strongest 980nm absorption line parallel to the optical $N_m$-axis (rotated by $\sim\,20^\circ$ from the crystallographic $a$-axis\cite{Biswal_thermo_optical_05,mochalov_laser_1997,pujol_crystalline_1999}).
The laser head design is based on the thermal lens shaping (TLS) technique developed in-house for astigmatism compensation in diode-pumped Nd-doped lasers with Brewster-cut gain media\cite{Rimington_thermal_lens_2004}.
It employs a simple single layer \ce{SiO2} anti-reflection coating on the gain medium to provide minimal loss for $p$-polarized intracavity 1040nm laser radiation and less than 0.5\% surface reflection for orthogonal $s$-polarized 980nm pump light.
The result is a simple and robust, soliton mode-locked and directly diode-pumped, solid-state laser oscillator delivering over 4W of average output power in a 63MHz train of pulses with a duration less than 300fs.
Frequency doubling of this laser output in a 2mm Brewster-cut Lithium triborate (LBO) crystal provides 1.65W of average power at 520nm.

The Watt-level green output power from the frequency-doubled sub-picosecond Yb:KGW oscillator corresponds to a visible peak laser pulse power in excess of 100kW, which is well-suited for further frequency conversion through harmonic generation or pumping of an ultrashort pulse optical parametric oscillator.
%TODO change voice/check ->
To our knowledge, the generation of subpicosecond green peak pulse powers greater than 100kW by frequency doubling the output of a laser oscillator has only been demonstrated for high-power thin-disk\cite{Brunner_powerful_2004,Marchese_pulse_energy_06} and cavity-dumped\cite{Palmer_microjoule_07} Yb-doped solid-state lasers, although externally-doubled commercial Ytterbium femtosecond oscillators\cite{website_amplitude,website_solar,website_high_q,website_time_bandwidth} and oscillators with rod-like Yb-fiber gain media [22] can be expected to yield comparable peak green pulse powers.
%TODO replace to "This output will then be use to ..." and link to something from previous
%In our case, the visible femtosecond pulses will be used as the driving laser radiation source for an ultrafast electron microscope [23] – a technology that seeks to combine the sub-nanometer spatial resolution of electron microscopy with the sub-picosecond temporal resolution afforded by today’s ultrashort pulse laser systems.
%Critical to this technology is likely to be the development of a femtosecond photo-electron gun with a low spatial emittance [25].
%The coherent control of photoemission through visible laser-driven plasmon excitations on large-area (a few mm2), gold or silver, nano-patterned photocathodes may provide a mechanism by which this can be achieved [25].

@@

The design of the laser head follows the example of Holtom\cite{Holtom_mode_locked_2006}, which ensures efficient pumping and access to the broadest emission bandwidth of Yb:KGW, while also employing the thermal lens shaping (TLS) technique to compensate for astigmatism\cite{Rimington_thermal_lens_2004}.
To ensure efficient operation of a Yb:KGW laser, the dominant absorption feature at 980nm for radiation polarized parallel to the optical $N_m$-axis ($\sim\,20^\circ$ from the crystallographic a-axis) should be pumped using high brightness diode lasers.
On the other hand, as with \ce{Yb$:$KY(WO4)2} \cite{Liu_diode_pumped_2001,Killi_high_peak_2005}, the broadest emission bandwidth for ultrashort pulse generation occurs for radiation polarized parallel to the optical $N_p$-axis (parallel to the crystallographic $b$-axis\cite{Holtom_mode_locked_2006}) in this biaxial material. 
%page 8632
Instead of using a polarization-coupled pumping geometry\cite{Holtom_mode_locked_2006} to ensure these optimum pump-lasing conditions, we employ the non-Brewster crystal geometry and cut shown in Fig. 1(a), where a single 193nm-thick \ce{SiO2} anti-reflection coating is applied to both polished 3x10mm Yb:KGW crystal faces.
At an angle of incidence of 32.3$^\circ$, this coating theoretically has zero reflection loss for $p$-polarized 1040nm laser radiation parallel to the $b$-axis propagating perpendicular to the $a$-$b$ plane of the crystal (Fig. 1(b)); i.e., the preferred optical $N_p$-axis with refractive index $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}.
The \ce{SiO2} coating also has an appropriate bandwidth for femtosecond laser operation in excess of 100nm (to the
0.1\% reflection points) for $p$-polarized radiation centered at 1040nm.
On the other hand, the $\sim$ 0.4\% reflection loss per surface for $s$-polarized 1040nm radiation is sufficient to suppress $s$-polarized oscillation in a laser resonator.
In addition, the same anti-reflection coating has less than 0.5\% reflection loss for $s$-polarized 980nm pump radiation (polarized parallel to the $a$-axis of the Yb:KGW crystal) at incidence angles between 21 and 39$^\circ$ (Fig. 1(b)).
Thus, in this Yb:KGW laser crystal geometry, the largest absorption cross-section associated with the optical $N_m$-axis is accessed with about 94\% efficiency ($\cos 20^\circ = 0.94$) by the $s$-polarized pump radiation.

@@

To ensure a stigmatic laser resonator, we also employ the thermal lens shaping (TLS) technique\cite{Rimington_thermal_lens_2004} to compensate for the astigmatism induced by the non-normal incidence geometry of the laser crystal. 
For the employed non-Brewster crystal cut (Fig. 1(a)), a ray
%page 8633
transfer matrix analysis (with the $x$ direction parallel to the crystallographic $b$-axis and the $y$
direction parallel to the $a$-axis) reveals that the required ellipticity for a stigmatic thermal lens
generated by a parabolic temperature distribution, $T(x, y) = T_0 - \frac{1}{2} ( A_\smallT x^2 + B_\smallT y^2 )$, where $T_0$
is the on-axis temperature, is given by
%TODO check W symbol consistent with the rest of the paper
\begin{equation} \label{eq:laser_ratio_ba}
  \frac{B_\smallT}{A_\smallT} = 
  \frac{ \cos^2 \theta_2 \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) }
       { \cos^2 \theta_1 \left[ \left( \dfrac{dn}{dT} \right) + 2 \Delta \alpha_\smallT \left( n \cos\theta_2 - \cos\theta_1 \right) \right] }
  \quad\text{.}
\end{equation}
Here, $\frac{dn}{dT}$ and $n$ are the thermo-optic coefficient and refractive index of the medium respectively (in this case for radiation polarized along the optical $N_p$-axis), $\theta_1$ is the 32.3$^\circ$ angle of incidence, $\theta_2$ = 15.7$^\circ$ is the angle of refraction (as $n_p$ = 1.98 \cite{Biswal_thermo_optical_05,pujol_crystalline_1999}), and $\Delta$ is the fraction of the crystal length $l$ contributing to the bowing of the crystal faces under thermal
expansion with coefficient $\alpha T$ \cite{Rimington_thermal_lens_2004}.
Generally, $\Delta \approx W l$ , where $W$ is the transverse size (radius) of the pumped region [30].
In our case, $l \approx$ 3mm and $W \approx$ 0.2mm, so that $\Delta < 0.1$, implying that the thermal duct due to a non-zero $\frac{dn}{dT}$ primarily determines the required $B_T$/$A_T$ ratio.
In other words, for the potassium gadolinium tungstate (\ce{KGd(WO4)2}) crystal geometry of Fig. 1(a), we expect from \ref{eq:laser_ratio_ba} to require $\dfrac{B_\smallT}{A_\smallT} \approx \dfrac{ \cos^2 \theta_2 }{ \cos^2 \theta_1 } = 1.3$.

The spatial pump distribution needed to generate the required $B_T/A_T$ ratio can be found using a cosine Fourier series solution to the thermal diffusion equation for our two-dimensional problem,
\begin{equation} \label{eq:heat_eqn}
  \kappa \nabla^2 T(x,y) = Q(x,y)
  \quad\text{,}
\end{equation}
where $\kappa$ is the thermal diffusion coefficient.
Thermal diffusion in the third $z$-direction has been neglected due to the relatively uniform longitudinal heat deposition resulting from our counter-propagating pump geometry\cite{Rimington_thermal_lens_2004}.
The specifications of our dual-axis collimated diodes [26] indicate that heat source can be written as
\begin{equation}
  Q(x,y) = Q_0 \exp \left[ - \frac{x^2}{W_x^2} - \left( \frac{y^2}{W_y^2} \right)^3 \right]
  \quad\text{;}
\end{equation}
that is, the focused pump radiation is described accurately by a Gaussian irradiance distribution in the horizontal ($x$) direction with half-width 1/e maximum (HW1/eM) spot size $W_x$ and a super-Gaussian of order 3 in the vertical ($y$) direction with spot size $W_y$.
\ref{eq:heat_eqn} can then be solved using the appropriate boundary conditions for our 3x10mm crystal cross-section for the employed TLS technique\cite{Rimington_thermal_lens_2004}; namely, heat removal only through the top crystal face implies $T(x,y = \pm 1.5\text{mm}) = 0$ (or arbitrary constant) and $\dfrac{\partial T}{\partial x} = 0$ at $ x = \pm 5.0$mm for any $y$.
For a fixed vertical pump spot size of $W_y = 0.2$mm, our analysis indicates that in order to achieve the required ratio $B_T / A_T = 1.3$ in the parabolic approximation for $T(x,y)$ about $(x,y) = (0,0)$ a horizontal pump spot size $W_x = 0.3$mm is needed; in other words, the required $Q(x,y)$ heat source spot size ratio (or ellipticity) $W_x / W_y = 1.5$.
This analytical result for the central region around the laser beam axis is shown in Fig. 2; displayed are the
%page 8634
horizontal ($y = 0$, Fig. 2(a)) and vertical ($x = 0$, Fig. 2(b)) sections, respectively, through the $Q(x,y)$ and $T(x,y)$ distributions.
Also shown are the parabolic approximations to the pump-induced temperature distribution (dashed lines) and representative TEM$_{00}$ Gaussian laser modes (shaded area) in the Yb:KGW crystal with a HW1/eM field mode sizes of $w_y = 160\mu$m and $w_x = 182 \mu$m (14\% larger due to refraction (Fig. 1(a)).
This mode size ensures efficient pump-probe overlap while minimizing the loss due to ground state absorption in the unpumped regions of the Yb-doped gain medium\cite{Brenier_new_criteria}; in particular, in the vertical direction (Fig. 2(b)) where $w_y$ greater than about $0.8 W_y$ leads to significant absorptive loss as the extrema of the laser mode extend beyond the sharply-bounded super-Gaussian pump region.

@@

We note that the above thermal lens analysis neglects the anisotropy of the thermal conductivity $\kappa$ in potassium gadolinium tungstate (\ce{KGd(WO4)2}) \cite{Biswal_thermo_optical_05,mochalov_laser_1997} and stress-induced refractive index changes [31].
For the employed crystal cut in our TLS geometry (Fig. 1(a)), the anisotropy in $\kappa$ \cite{mochalov_laser_1997} [27] is not expected to have a significant effect; the laser head geometry dictates that the dominant pump-induced heat conduction will be in the saggital ($y$) direction along the crystal $a$-axis where $\kappa \approx 2.6$W/(mK) (heat extraction only through the top and bottom 3x10mm crystal faces), so that the 46\% larger value of $\kappa$ along the $b$-axis ($x$) direction at 3.8W/(mK) should not have a large influence on the heat conduction.
The refractive index
%page 8635
changes due to stress in the pumped gain medium are difficult to quantify since the substantial variations in the reported values of the thermo-optic coefficients in KGW make it impossible to extract either the sign or magnitude of the stress dependence of the refractive index\cite{website_time_bandwidth}.
Moreover, nearly all analyses of thermal lensing effects Nd:KGW and Yb:KGW lasers have been performed for laser radiation polarized along the optical $N_m$-axis, which accesses the largest emission cross-section for both Nd- and Yb-doping [31,34].
Consequently, for our laser polarized along the optical $N_p$-axis\cite{Holtom_mode_locked_2006}, there is insufficient information to include accurately the effects of stress-induced refractive index changes.
